Finite element analysis ofmulti-dimensional and simplifiedmodels for beams and plates

### 2023-12-03 files/Overview.tex
summary

\section{Achievements}

In the first chapter, the models of this dissertation are presented and a dimensionless variational form for each is derived. The model problems for this dissertation are then formulated.

In the second chapter, two results are investigated. This first is on the existence and uniqueness of solutions for vibration problems. An article is discussed that proves the existence and uniqueness of solutions for a general vibration problem using semi-group theory and assuming four assumptions. All the models in this dissertation is a special case of this general problem, so the theory can be explained and applied to an example. The weak variational form for a cantilever Timoshenko beam is derived. In this derivation, we also give the definitions of the spaces required. The theory is then applied to this example problem. We show that the four assumptions can be proven to be true for the cantilever Timoshenko beam model.

The next result is in Modal Analysis. Again the cantilever Timoshenko beam model is used to first explain the theory, before the general case is discussed. A trial solution is presented and substituted into the partial differential equations. This results into two problems, an eigenvalue problem and an ordinary differential equation. These two problems are solved and substitution can confirm then that the trial solutions are indeed solutions. The method obtains the same result as separation of variables, but avoids any division. Finally a formal series solution for the example is given. The general case is then discussed and follows the same idea. Modal analysis is crucial for this dissertation. For the comparisons later in this dissertation, a method is required to compare the different models. Modal analysis shows that is the eigenvalues and eigenfunctions of the models compare well, then the solutions of the models should compare well. And since the Fourier Coefficients gets small very fast, it is only required to look at the first few eigenvalues.

In the third chapter, two theoretical results for the Finite Element Method is discussed. The first result is on the convergence of the Galerkin Approximation. The general case contains various symbols rectangularly used in application, so an example of the cantilever Timoshenko beam model is used and the Galerkin Approximation is derived. Then the general case is discussed from an article. The results from the article are presented in a reduced format to avoid introducing too many new symbols. We also present the results in four new Theorems, to present the results of the article in a concise manner.

The second result is on the convergence of the eigenvalues and eigenfunctions when using the Finite Element Method. The results are presented in textbook. Here we use the notation of the dissertation to present the results. The results are also expanded in an attempt to better explain the theory, with a few new parameters also introduced.

In the fourth chapter, we investigate our simplest model, the Timoshenko beam theory. First is a discussion of an article that provides a method that allows us to find the exact eigenvalues and eigenfunctions of a Timoshenko beam model. A summary of this method is presented. It is then applied to two model problems for a cantilever Timoshenko beam and a free-free Timoshenko beam. For both model problems, the method is applied and examples of the eigenvalues and mode shapes are given. The eigenvalues and mode shape plots are obtained using a computer program, sketching the equation of motion and the using interval division to get the eigenvalues to a desired level of accuracy. This is in preparation for the comparison results in the results that follow.

In the fourth chapter, we look at our simplest model, the Timoshenko beam model. The first result discusses an article that provides a method to determine the exact eigenvalues and eigenfunctions for the Timoshenko beam model. We then apply this to a cantilever beam model and a free-free beam model to obtain the eigenvalues and plot the mode shapes. We then look at an article that compares a cantilever Timoshenko beam model to a two-dimensional beam model. We discuss the results of the article and replicate the results with more significant digits. We also include some more results not included in the article, that are used later in Chapter 6. We then look at another article that compares the Timoshenko beam theory to results obtained from a measured beam. The authors suspended a beam on both ends an vibrated the beam until then endpoints could be considered free. They then swept through different frequencies and recorded the natural frequencies of the beam. These results are then compared to the theoretical results of a free-free Timoshenko beam. We add to the article by giving the model problems for the beams and discuss the results.

The fifth chapter is a preparation chapter for the comparisons of this dissertation. In chapter 4, we have a method to obtain the exact eigenvalues and eigenfunctions for the Timoshenko beam theory. But a different approach is required for the two- and three-dimensional models. In this chapter, we derive the Galerkin Approximation for each of the models. We then define a discrete grid of elements on the reference configuration on each of the models. Then using piece-wise Hermite cubic basis functions, we derive a matrix representation of the model problems using the standard Finite Element Analysis. In this form it is then easy to obtain a matrix representation of the eigenvalue problem for each of the models. This is in preparation for the comparisons of the models in Section 4.5 as well as chapter 6.

In the final chapter, we extend the work of Section 4.5 and investigate the validity of the two-dimensional beam model as well as the validity of the Reissner-Mindlin plate model. Using the work from chapter 5, a numerical formula is derived to calculate the eigenvalues and eigenvectors of the models. This is implemented into some MATLAB code. (This was also done to obtain the results in Section 4.5). The same structure as Section 4.5 is followed. To start, the mode shapes are calculated and plotted. These are then compared so that the eigenvalues can be matched up. Careful consideration should be taken to the shapes as there are many shapes not relating to beam-type or plate-type problems. The results show that the shape of the models impact how well the models compare. The cases of the comparisons were chosen so that they represent realistic scenarios for use in application.

\section{Future Work}

>>>>

\section{Contributions}

There are four models used in this dissertation. For each of the models, the dimensionless variational form is derived. Also presented are the model problems that are used in this dissertation.

The first result investigates the existence and uniqueness of solutions for general vibration problems. The article that is discussed proves this result for a general vibration problem, using four assumptions. To explain the theory, an example is presented using one of the model problems of the dissertation. The cantilever Timoshenko beam model is chosen for the simple boundary conditions, as well as it's recurring importance in this dissertation. The weak variational form of the model is derived as well as the function spaces defined. This is presented in the same format as the general vibration problem. In fact all the models in this dissertation are special cases of this general vibration problem. The theory is then applied to this example problem, as a demonstration. To apply the theory, the four assumptions are proven to be true.

The next result looks at modal analysis. Before the general case is discussed, again the cantilever Timoshenko beam is used to illustrated the concept of modal analysis.A trial solution to the boundary value problem is suggested. This trial solution is substituted into the partial differential equation and two problems are obtained. The first problem is the eigenvalue problem and the second is an ordinary differential equation. The eigenvalue problem can be solved with theory discussed later in the dissertation and the ordinary differential equation can then be solved. Substation of these two results into the boundary value problem confirms that the trial solution is correct. The same idea is then discussed for the general case.

We then look at the convergence of the Galerkin Approximation for our general vibration problem. The results of the article discussed are updated with improved notation using a different article. The general case of the Galerkin approximation use a lot of symbols that are not immediately obvious, so again the cantilever Timoshenko beam model problem is used and the Galerkin Approximation is derived in an attempt to explain some of the conventions. The results of the article are then discussed and presented. The results are presented in a concise and practical format to reduce the need to define any unnecessary symbols or notation. The results are also presented in four theorems, summarizing the results of the article that are important but that are not necessarily presented as a theorem in the article.

The next result is on the convergence of the eigenvalues and eigenfunctions of a general vibration problem when using the Finite Element Method. The results are from a textbook. The results are presented with updated notation, coinciding with the notation used in the dissertation. The results are also expanded and extra results are added in an attempt to better explain the theory.

We then look at an important result for the Timoshenko beam theory. This provides a method to calculate the exact eigenvalues and eigenfunctions for the Timoshenko beam theory. Two examples are then used, a cantilever beam and a free-free beam, as an example of the application of the theory. The eigenvalues are calculated and the corresponding mode shapes are plotted. To obtain these results, the equation of motion is plotted, the isolation intervals are determined and the eigenvalues are then calculated using interval division to a desired level of accuracy. The mode shapes can then also be plotted with back substitution. These examples are important preparation for the main comparisons of this dissertation. We also discuss a result comparing the Timoshenko beam theory to results from an empirical study. We add to this article by giving the model problems.

For the rest of the models in this dissertation, the two- and three-dimensional elastic bodies, and the Reissner-Mindlin plate model, a different approach is required to solve the eigenvalue problems. For these models we use the Finite Element Method. For each of the models, their reference configurations are given. All of the models are assumed to have a square cross-section and in a cantilever configuration. This reference configuration is then discretised into a grid of rectangular shaped elements. A set of admissable piecewise Hermite cubic functions are then used and each model is rewritten into a Galerkin Approximation. We the define the standard Finite Element Method Matrices for each case. These are referred to as the mass and stiffness matrices in engineering. Finally our boundary value problems are written into a system of ordinary differential equations in a matrix representation. At this point is it easy to derive the eigenvalue problem for each of the problems in this matrix form. This is in preparation for the main comparisons made in this dissertation.

For our main comparisons, we first look at an article comparing a cantilever Timoshenko beam model to a cantilever two-dimensional model. We discuss the article and replicate the results. The eigenvalues and eigenfunctions of the Timoshenko beam model is obtained using the exact method already described. The eigenvalues and the eigenfunctions for the two-dimensional model are approximated using the Finite Element Method matrix representation of the eigenvalue problem. A MATLAB program is written to approximate the eigenvalues and plot the mode shapes. The accuracy of this approximation is also investigated by looking at the rate of convergence for different grid sizes. Following the article, the eigenvalues are matched up by first comparing the mode shapes of the two models. The eigenvalues can then be matched up and also any eigenvalues not relating to beam-type problems can be filtered out. Based on the results in modal analysis, only a few eigenvalues need to be considered. The relative error between the eigenvalues are then calculated for different shapes of beams. The results improve on the article by showing more significant digits and also including some more results.

We then extend the results of the article to investigate the validity of a cantilever two-dimensional beam model as well as a cantilever Reissner-Mindlin plate model. The same method of the article is followed. To investigate the validity of a cantilever two-dimensional model, a cantilever three-dimensional model is considered. Since both of the models are not beam models, careful consideration needs to be taken to identify the eigenvalues. The same approach is used by comparing the mode shapes. For interest, the non-beam type eigenvalues shared between the two models are also included and the rest that the three-dimensional model does not share with the two-dimensional model are omitted. A clear distinction is made to show the beam type eigenvalues and the beam type results. The shape of the models are also carefully chosen to represent a variety of realistic cases that are interesting. The results show that the shape of the models play an important role in how well the models compare. An interesting result shows that this comparison is not good when the beam gets too wide. We therefore suggest a different model, like a plate model. This lead to the introduction of the Reissner-Mindlin plate model into this dissertation. The validity of a cantilever Reissner-Mindlin plate model is investigated using the same method. The cantilever three-dimensional plate model is again used as the reference model with the restriction that the body is wide.

\section{Further Research}
Future work would include the addition of damping into the models. A lot of the articles do include results for damping, however the results of the modal analysis (the crucial theory of this dissertation) might not be so trivial to prove.

The use of the Hermite cubic basis functions results in the derivatives of the displacement functions to also be available. This brings into question if the stresses of the models can also be compared.

Further improvements to the code can be made. Although a lot of effort was put into optimizing the code, this lead to the code being difficult to understand. Ideally a refactor and simplification of the code, while maintaining its functionality is desired.

### 2023-12-03 summary.tex
summary

\begin{center}
	\large \textbf{Summary}
\end{center}
This dissertation is a literature study, investigating the Finite Element Analysis of multidimensional and simplified beam and plate models. The Timoshenko beam and Reissner-Mindlin plate theories are popular theories used in mathematics and engineering. An example of ongoing research using these theories is the study of suspension bridges. The articles \cite{Mck99}, \cite{BOC20} are examples of mathematical studies while \cite{SV86} follows an engineering approach.\\

 This dissertation is a literature study on the Finite Element Analysis 
 \begin{itemize}
	\item Considering linear mathematical model s for elastic beams and plates.
	\item Focus is on the one-dimensional Timoshenko beam theory and two-dimensional Reissner-Mindlin plate theory
	\item Also require two and three-dimensional beam models and a three-dimensional plate model
	\item Models are used in variational form (Chapter 1)
	\item Existence and uniqueness of solutions (Chapter 2)
	\item Convergence of FEM (Chapter 3)
	\item Chapter 4: - Timoshenko beam theory - Modal analysis - Examples of calculating eigenvalues and eigenvectors - Emperical study - Tipbody model with numerical results.
	\item Chapter 5: FEM applied to cantilever 3D, 2D and plate elastic bodies
	\item Chapter 6: Validity of Timo and Reissner-Mindlin Models
\end{itemize}

\end{document}

>>>>

\begin{center}
	\large \textbf{Summary}
\end{center}

This dissertation is a literature study that investigates the validity of different linear models in application. Validity in this context refers to how well simplified lower-dimensional models, such as the Timoshenko beam and Reissner-Mindlin plate models, compare to more realistic higher-dimensional models, including a two-dimensional beam, and three-dimensional beam and plate models. The models in this dissertation are all special cases of a general vibration problem.

First, the dissertation examines the existence and uniqueness of the general vibration problem. An example is used to to explain the theory, which is then subsequently applied to by proving that the assumptions are satisfied.

Following this, the concept of modal analysis is introduced using an example, before delving into the general case. These results on modal analysis are crucial to the dissertation, as they explain that the solutions of the models will compare well if the eigenvalues and eigenfunctions of the models compare well.

The dissertation then explores two theoretical results for the Finite Element Method (FEM). The initial result involves an analysis of an article on the convergence of the Galerkin Approximation. The findings of the article are reformulated as theorems with simplified notation for clearer presentation

Subsequently, the dissertation reviews results from a textbook regarding the convergence of eigenvalues and eigenfunctions in a general vibration problem when utilizing FEM. These results are adapted with updated notation and expanded upon for a more comprehensive explanation of the theory.

Concerning the Timoshenko beam model, the dissertation investigates an article that presents a method to calculate the exact solutions of the eigenvalue problem. Two practical examples are provided to illustrate the application of this method. Additionally, the dissertation looks at an article that compares the theoretical results of the eigenvalue problem with an empirical study done by the authors.

For the remaining models, the dissertation employs FEM to solve the eigenvalue problems. The boundary value problems of each model is rewritten as systems of ordinary differential equations in matrix form using FEM. The eigenvalue problems are then derived from this matrix representation. Piecewise Hermite cubic basis functions are used, and the solutions of the eigenvalue problems are approximated using MATLAB scripts.

In investigating the validity of simplified models, the dissertation first considers an article comparing the Timoshenko beam model to a two-dimensional beam model. The authors method of comparison is discussed, and their results are replicated with a higher degree of accuracy. The dissertation then extends this approach to assess the validity of a two-dimensional beam model and a Reissner-Mindlin plate model, using the three-dimensional model as reference. The method to compare the models are the same as in the article. First the eigenvalues are calculated, sorted and matched by analyzing the corresponding mode shapes. The mode shapes are also used to identify eigenvalues specific to beam- and plate-type problems. The error can then be calculated. Different shapes of beams and plate models are considered that are realistic in application.

The derivation and comparison of the two and three-dimensional models is the main contribution of this dissertation.
\end{document}
